Feeding Dynamic Behavior into Optimization: Simulation-based Multi-objective Overtime Planning

### 2014-08-28 data/result/Article/Latex/Article.tex
Nova versão do artigo

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}

\usepackage{bigstrut}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{booktabs}
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\author{\IEEEauthorblockN{Márcio de O. Barros\IEEEauthorrefmark{1},
						  Luiz Antonio O. de Araujo Jr\IEEEauthorrefmark{2}
        \IEEEauthorblockA{Federal University of the State of Rio de Janeiro\\
Av. Pasteur 458, CEP 22.290-240
Rio de Janeiro, Brazil
\\ \{marcio.barros\IEEEauthorrefmark{1}, luiz.araujojr\IEEEauthorrefmark{2}\}@uniriotec.br}
}}

\begin{document}
\maketitle
\begin{abstract}
IT professionals are particularly subjected to working overtime due to difficulties in estimation, requirement specification, and measuring progress during a software development life-cycle. However, excessive overtime has detrimental effects on developers and the software they produce. This contrast creates the need for overtime-planning strategies to support project managers in taking the most from as few overtime hours as possible. In this paper we introduce a formulation for the overtime-planning problem which extends the state-of-art by taking into consideration both the positive effects of overtime on productivity and its negative effects on product quality. We use heuristic optimization to explore close to optimal overtime allocations under this formulation. We present an empirical study that compared our approach with current industrial practices and a similar formulation without negative effects. Results show that suppressing the flip-side of the gains brought by overtime may lead to wrong decisions. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan by 5\% and 3\%, respectively, due to longer testing activities required to fix the exceeding errors introduced by overtime. Evidence also supports an overtime-management strategy used in the industry that concentrates overtime in the second half of a project schedule to compensate for early delays - about 60\% of the overtime allocations produced under this strategy are optimal.
\end{abstract}

\begin{keywords}
Overtime planning; project management; simulation; schedule optimization.
\end{keywords}

@@

This paper extends the state-of-the-art by taking into account the documented impact of overtime on the quality of the services provided by developers working in extended hours. It introduces a new formulation for the overtime planning problem which combines search-based optimization with a scenario-based simulation that takes into account an increase in the number of errors injected into a software product by a team working overtime. We compared our results to current practices on regard of overtime management and provide insights to support project managers on deciding when to use overtime work. By doing this, we found scientific support for an overtime-management practice used in the industry, that is, evidence aligned with project management intuition. Our results show that, on average, 60\% of the allocations that might be built according to this strategy are optimal. We also compared our results with a similar model that ignores the dynamics of error generation caused by overtime,  showing that wrong decisions may be taken if the increase in error generation rates is not taken into account. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan, respectively, by 5\% and 3\% due to longer testing activities required to fix the exceeding errors introduced by overtime. The primary contributions of this paper are:
\begin{enumerate}
\item We introduce a new formulation for the multi-objective overtime-planning problem that considers an increase in error generation rates due to fatigue. Our approach takes into account trade-offs among the number of overtime hours worked in a project, makespan, and cost;

\item We present an empirical study based on 6 real-world projects with up to 635 function-points in size. Results show that concentrating overtime hours in the second part of project schedules (a practice used in the industry) is a good strategy that can be improved by optimization;
\item We present evidence that the negative impact on product quality caused by overtime must be taken into account while selecting the overtime allocation for a project. A strategy that may be optimal if such impact is disregarded may not be the best choice if it is considered.
\end{enumerate}

This paper is divided into eight sections, starting with this introduction. Section II presents our formulation for the overtime-planning problem, as well as the simulator used to capture the dynamics of error propagation throughout the software development process. Section III presents the proposed solution for the overtime-planning problem. Section IV presents the design of the experimental analysis carried out to evaluate the proposed solution. Section V presents the analysis and results collected from the experimental study. Section VI addresses validity threats that may limit the generalization of experimental results. Section VII presents related work and Section VIII shows conclusions and future research directions regarding overtime planning.

@@

We depart from a set of work packages, each to be analyzed, designed, coded, and tested. A project is represented as an acyclic directed graph $<$WP, WPDEP$>$ consisting of a node set WP = \{wp1, wp2, ..., wpn\} of work packages and an edge set WPDEP = \{(wpi, wpj): i$\not=$j,1$\leq$i$\leq$n,1$\leq$j$\leq$n\} of precedence dependencies amongst work packages. Each wpi is described by the amount of effort required to elicit its requirements, the amount of effort to design it, code it, and test it as if these tasks were performed by a mean developer in a regular work environment. A work package is also described by the number of errors its requirements analysis, design, and coding activities might generate in a similar context. On Section II.B we discuss how these parameters were filled for our instances. Each dependence in WPDEP represents a work package (wpi) whose development can only start after the requirements for a second work package (wpj) are known.

From these work packages, we develop a project schedule. Our schedule is a partial order of software development activities on which four types of activities are created for each work package: requirement analysis, design, coding, and testing. A testing activity is always preceded by a coding activity, which is preceded by a design activity, which comes after a requirement analysis activity. These dependencies represent the basic workflow of modeling the problem, designing the solution, implementing its source-code, and testing its features. The schedule also includes dependencies amongst work packages: the requirements analysis activity for work package wi cannot start before the requirements analysis activity for wj is concluded for all (wi, wj) pairs in WPDEP.

A project schedule is represented as an acyclic directed graph $<$ACT, ACDEP$>$ consisting of a node set ACT = \{a1, a2, ..., an\} of activities and an edge set ACDEP = \{(ai, aj): i$\not=$j,1$\leq$i$\leq$m,1$\leq$j$\leq$m\} of finish-start (FS) dependencies amongst activities. A FS dependency A$\rightarrow$B indicates that work on activity B can only start when activity A is finished. Each activity ai is classified as either requirement analysis, design, coding, or testing. Analysis, design, and coding activities are described by the expected effort required to produce its results and the expected number of errors they might generate if performed by a mean developer in a regular environment. Testing activities are described by the average effort required to identify and correct an error, which is collected from information provided in work packages.
We consider three objectives for overtime planning: the number of overtime hours (NOH), project makespan (MKS), and overall project cost (CST). OPP is formulated as a three objective decision problem in which makespan, cost, and overtime hours are conflicting objectives to be minimized. A candidate solution for the OPP is a sequence of numbers that represents the amount of daily overtime hours to be spent on each activity comprising the schedule. Our formulation is based on a 40-hours work week, accepts multiples of 30 minutes for overtime work, and limits the maximum amount of daily work to 12-hours turns, according to Brazilian law. Thus, for any given activity we may assign zero overtime, 30 minutes of overtime, one hour, one hour and 30 minutes, and so on up to four hours of daily overtime. Our first objective, NOH, is calculated as the sum of the amount of overtime hours assigned for each activity in the schedule.

The $<$ACT, ACDEP$>$ graph is comprised of a set of paths $\Pi$ denoting the precedence-respecting sequences of activities that must be executed in order to complete the project. Based on these paths, the makespan of a project (MKS) is calculated by Equation (1). The duration of a given activity is calculated by a simulator according to the characteristics of the activity and its execution setup (including the number of daily overtime hours assigned to it). The simulator calculates the number of errors produced by each development activity and propagates these errors to succeeding activities. For testing activities, the simulator calculates the effort required to identify and correct all errors received from former activities. The makespan of a project is given by the longest path in $\Pi$, which is known as the critical path.

\[
MKS = MAX_{p\in\Pi} (\sum_{a_{i} \in p} Duration (a_{j}))
\]

@@

\subsection{Simulation Machanism}
Simulation is the imitation of the behavior presented by a real-world process or system over time \cite{Banks:2000}. It supports predicting and understanding the behavior of complex systems, containing hundreds or thousands of interconnected components. Simulation is useful because the number of components and the complexity of the data and control exchange among them usually hampers our ability to interpret system behavior by intuition or analytical reasoning.

@@

To overcome some of these problems, we have designed an object-oriented, continuous-time simulator to model the effects of overworking in software projects and predict project behavior (cost and makespan) accordingly. The simulator is based on three constructs: simulation objects, resources, and scenarios. Instead of creating a specific syntax to represent these elements, we relied on the Java programming language to describe and execute them. Designing the simulator in Java also simplified its integration to the heuristic search framework we have used in our experimental study.
Simulation objects are the elements whose behavior we want to simulate. They represent the nodes of a directed and acyclic graph whose edges denote precedence dependencies amongst simulation objects. They are initialized before the simulation starts and thereafter kept dormant until all their dependencies have been processed. Then, they become alive and start participating in the simulation until signaling their conclusion to the simulator. Simulation ends when all objects presented to the simulator are finished.
Time is managed in a continuous fashion. Simulation starts at time zero and is comprised of consecutive simulation steps, each advancing the simulator's clock by an infinitesimal increment. At each step, all alive simulation objects are allowed to update their internal state. The internal state of an object is defined according to the relevant characteristics of the real-world element it represents in the simulation. The behavior of an object is thus represented by the changes observed in its internal state during the simulation.
Simulation objects are Java objects that implement an interface defining a protocol through which they are controlled by the simulator. The internal state of an object is kept in its instance variables. Internal state adjustment operations are coded in methods activated by the simulator: init(), to initialize the simulation object at time zero; start(), called when all dependencies for an object are finished; step(), called at each simulation step; and finish(), called when the object signals its conclusion.
During a simulation step, simulation objects may use resources. A resource has a limited amount of materials or capabilities that can be consumed by simulation objects. These materials or capabilities are refilled before each simulation step. An object may request part of the resource's stock and adjust its internal state accordingly to its availability. The simulator controls the usage of limited resources by different simulation objects running in parallel. To model a software project, simulation objects represent an activity network with FS dependencies. The developers available to work on activities are resources whose amount of work is limited to a certain number of hours per day and is renewed in a daily basis. Developers are assigned to activities and if two or more activities executing in parallel share the same developer, only one of them will use the developer's capabilities at a time.
Scenarios are used to represent complementary behavior that may be imposed to certain simulation objects. A scenario is written to influence a certain type of simulation object and only participates in the simulation if activated on one or more objects of that type. Scenarios can query and update the internal state of their associated objects before and after each event on which the objects update their internal state (start, step, finish). Scenarios can be used to represent managerial decisions imposed to a subset of a software project's activities. They are reusable model components that modify the behavior of their related simulation object according to the policy they represent. For the purpose of the OPP, a scenario was developed to represent the effects of overtime while a base model describes the expected behavior for a project, its activities and developers as if uninfluenced by managerial policies and decisions (such as overtime allocation).

\subsection{The Base Model}

The base simulation model is comprised of activities and developers. Activities are simulation objects linked to each other by precedence dependencies. An activity is available for simulation as soon as its precedent activities are concluded. Developers are resources who can spend a limited amount of work on each simulation step. They are characterized by their productivity and error generation rates.
Activities are separated into development and testing activities. Development activities are described by an expected amount of work to be performed (in developer-hours) and an amount of errors expected to be generated while this work is performed. The duration of a given activity depends on the amount of work to be performed and the productivity of the developer assigned to it. We calibrated the model using data from Jones \cite{Jones:2000}. Given a work package measured in function-points (FP), Jones \cite{Jones:2000} presents average distributions of effort required for each development phase. For instance, for projects of about 1,000 function-points, requirement analysis activities consume 7\% of project effort, design activities consume about 10\%, coding activities consume about 30\%, and testing consumes 30\%. The remaining effort is related to management, documentation, and other activities. In our model, this effort is proportionally distributed among the analysis, design, coding, and testing activities. Given this distribution, the size of a work package, and an average productivity of 27.8 function-points per developer-month \cite{Jones:2000}, we calculate the expected duration for each activity.
Error dynamics is trickier. When an activity starts, it collects all errors produced by its former activities (if any) carrying them toward the end product. Abdel-Hamid and Madnick \cite{Abdel-Hamid:1991} present a model for error reproduction which states that errors inherited by an activity are not only forwarded to succeeding activities, but they contribute to the generation of new errors. Jones \cite{Jones:2000} suggests that the average number of defects introduced by each kind of activity is 1 error/FP in requirement analysis, 1.25 errors/FP in design, and 1.75 errors/FP in coding activities. Based on these data, our model assumes that each activity introduces 1 new error/FP, design activities regenerate 25\% of errors it receives from analysis, and coding regenerates 33\% of the errors received from design. These regeneration rates lead to the mark of 4.0 errors/FP on average after the implementation and are compatible with the former model \cite{Abdel-Hamid:1991} and source \cite{Jones:2000}.
Testing activities receive the results of development activities and aim at identifying and correcting the errors introduced in the product. These activities are characterized by the average effort required to identify and correct an error. Their duration depends on this parameter along with the assigned developer’s productivity. Given the amount of work expected to be invested in the testing phase \cite{Jones:2000} (i.e., 30\% of total effort for 1,000 FP projects) and the expected number of errors reaching this activity in a regular situation (4 errors/FP), we calculate the average time required to identify and correct each error and use this value on testing activities.

\subsection{A Scenario for Overwork}

The scenario used in our simulation describes the dynamics of working overtime and its influence on developer's productivity and error generation rates. If a developer is expected to work more than 8 hours per day in a given activity, his productivity will be increased proportionally to the extra number of work hours, reaching up to a 50\% increase if working to the upper limit of 12 hours per day. However, according to de Marco’s law \cite{demarco:1999}, a developer working more than the regular daily hours may produce software below the quality standard. The more hours a developer works, the more tired he becomes. Tired developers become less careful to the work they are performing and, consequently, they tend to introduce more errors in the software. We have used data from \cite{Abdel-Hamid:1991}  to estimate the increase in the number of errors produced according to the number of hours worked by a developer. Such data suggest that software quality tends to deteriorate if developers work more than 9 hours/day and that a developer working 12 hours/day commits 50% more errors than while working on a regular shift.

\section{Solution Approach}
Our solution uses the NSGAII algorithm for optimization along with the simulator described in Section II. NSGAII \cite{deb:2002} is a multi-objective genetic algorithm based on a ranking procedure which classifies candidate solutions according to their dominance. A candidate solution is described by the values calculated for the objectives of interest under that solution. In our case, these objectives are the number of overtime hours, project makespan, and project cost, all to be minimized. A solution A is said to dominate a solution B if the values of all objectives in A are no greater than the respective objectives in B and at least one objective is smaller in A than in B. Non-dominated solutions are assigned a rank of 1; solutions dominated only by non-dominated solutions are assigned a rank of 2; solutions dominated only by the former are assigned a rank of 3, and so on. The algorithm evolves a population over a number of generations, applying crossover, mutation, and selection upon candidate solutions. The selection process prioritizes low-ranking solutions and, when a subset of solutions having the same rank must be selected, a density measure allows selecting candidate solutions covering the search space as uniformly as possible. 
As any heuristic algorithm, NSGAII has components that must be specified for a given problem. Three components are discussed below, while population size and stop criteria were subject of analysis and are discussed in the next section.

\textbf{Representation}. A candidate solution for OPP is a sequence of m integer numbers, m being the number of activities. Each number in the solution pertains to the [0..8] integer interval and represents a multiple of 30-minutes overtime for a given activity. Thus, if the number 4 is assigned to activity A, A will receive 1.5 hours of overtime work. Each individual solution being evolved by NSGAII is evaluated through simulation. The simulator activates the overtime scenario upon each activity in the schedule according to the number of overtime hours assigned to that activity in the solution. Then, the simulator calculates the number of overtime hours, project makespan, and project cost for that solution.

\textbf{Crossover}. We reused the crossover operator proposed by Ferrucci et al. \cite{Ferrucci:2013}. This operator was adopted because it was shown to outperform the one-point crossover operator commonly used in NSGAII for their formulation for the overtime planning problem. Instead of exchanging genetic material of two parents P$_{1}$ and P$_{2}$ after a point of cut C $\in$ [1..m], the resulting offspring O$_{1}$ and O$_{2}$ produced by this operator are defined by the equations below (given a random variable p with uniform distribution). The operator is executed with 50\% probability; otherwise, the parents are copied to the next generation. Parents are selected by binary tournament.

\begin{equation}
O_{1}(g)=
\left \{
\begin{array}{ll}
P_{1}(g) & 1\leq g < C \\
max(P_{1}(g), P_{2}(g)) & C\leq g\leq m,p < 0,5 \\
min(P_{1}(g), P_{2}(g)) & C\leq g\leq m,p \geq 0,5 \\
\end{array}
\right.
\end{equation}

\begin{equation}
O_{2}(g)=
\left \{
\begin{array}{ll}
P_{2}(g)& 1\leq g < C \\
P_{1}(g)+P_{2}(g) & C\leq g\leq m\\
\end{array}
\right.
\end{equation}

\textbf{Mutation}. As in Ferrucci et al. \cite{Ferrucci:2013}, mutations are executed for all crossover offspring. Each gene of an offspring O is changed with 30\% probability, being assigned a new value in the [0..8] integer interval.
\section{Exeperimental Design}
This section explains the design of our empirical study, presenting research questions, the instances selected to address these questions, and the measurements used to compare solutions found under different configurations and algorithms.
\subsection{Research Questions}

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{array}
\usepackage{bigstrut}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{booktabs}
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\author{\IEEEauthorblockN{M{\'a}rcio de O. Barros\IEEEauthorrefmark{1},
						  Luiz Antonio O. de Araujo Jr\IEEEauthorrefmark{2}
        \IEEEauthorblockA{Federal University of the State of Rio de Janeiro\\
Av. Pasteur 458, CEP 22.290-240
Rio de Janeiro, Brazil
\\ \{marcio.barros\IEEEauthorrefmark{1}, luiz.araujojr\IEEEauthorrefmark{2}\}@uniriotec.br}
}}

\begin{document}

\setlength{\floatsep}{2pt}
\setlength{\textfloatsep}{0pt}
\setlength{\intextsep}{2pt}
\setlength{\belowcaptionskip}{0pt}
\setlength{\abovecaptionskip}{0pt}

\maketitle
\begin{abstract}
IT professionals are particularly subjected to working overtime due to difficulties in estimation, requirement specification, and measuring progress during a software development life-cycle. However, excessive overtime has detrimental effects on developers and the software they produce. This contrast creates the need for overtime-planning strategies to support project managers in taking the most from as few overtime hours as possible. In this paper we introduce a formulation for the overtime-planning problem which extends the state-of-art by taking into consideration both the positive effects of overtime on productivity and its negative effects on product quality. We use heuristic optimization to explore close to optimal overtime allocations under this formulation. We present an empirical study that compared our approach with current industrial practices and a similar formulation without negative effects. Results show that suppressing the flip-side of the gains brought by overtime may lead to wrong decisions. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan by 5\% and 3\%, respectively, due to longer testing activities required to fix the exceeding errors introduced by overtime. Evidence also supports an overtime-management strategy used in the industry that concentrates overtime in the second half of a project schedule to compensate for early delays - about 60\% of the overtime allocations produced under this strategy are optimal.
\end{abstract}

\begin{keywords}
Overtime planning; management; simulation.
\end{keywords}

@@

This paper extends the state-of-the-art by taking into account the documented impact of overtime on the quality of the services provided by developers working in extended hours. It introduces a new formulation for the overtime planning problem which combines search-based optimization with a scenario-based simulation that takes into account an increase in the number of errors injected into a software product by a team working overtime. We compared our results to current practices on regard of overtime management and provide insights to support project managers on deciding when to use overtime work. By doing this, we found scientific support for an overtime-management practice used in the industry, that is, evidence aligned with project management intuition. Our results show that, on average, 60\% of the allocations that might be built according to this strategy are optimal. We also compared our results with a similar model that ignores the dynamics of error generation caused by overtime,  showing that wrong decisions may be taken if the increase in error generation rates is not taken into account. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan, respectively, by 5\% and 3\% due to longer testing activities required to fix the exceeding errors introduced by overtime. The primary contributions of this paper are:

\begin{enumerate}
\item We introduce a new formulation for the multi-objective overtime-planning problem that considers an increase in error generation rates due to fatigue. Our approach takes into account trade-offs among the number of overtime hours worked in a project, makespan, and cost;

\item We present an empirical study based on 6 real-world projects with up to 635 function-points in size. Results show that concentrating overtime hours in the second part of project schedules (a practice used in the industry) is a good strategy that can be improved by optimization;

\item We present evidence that the negative impact on product quality caused by overtime must be taken into account while selecting the overtime allocation for a project. A strategy that may be optimal if such impact is disregarded may not be the best choice if it is considered.
\end{enumerate}

This paper is divided into 7 sections, starting with this introduction. Section II presents our formulation for the overtime-planning problem and the simulator used to capture software project dynamics. Section III presents the proposed solution for the overtime-planning problem. Section IV presents the design of the experimental study carried out to evaluate the proposed solution. Section V presents and discusses the results collected from this study. Section VI presents related work and Section VII shows conclusions and future research directions regarding overtime planning.

@@

We depart from a set of work packages to be analyzed, designed, coded, and tested. A project is represented as an acyclic directed graph $<$WP, WPDEP$>$ consisting of a node set WP=\{$wp_1, wp_2, ..., wp_n$\} of work packages and an edge set WPDEP=\{($wp_i, wp_j$): i$\not=$j, 1$\leq$i$\leq$n, 1$\leq$j$\leq$n\} of precedence dependencies amongst work packages. Each $wp_i$ is described by the amount of effort required to elicit its requirements, the amount of effort to design it, code it, and test it as if these tasks were performed by a mean developer in a regular work environment. A work package is also described by the number of errors its requirements analysis, design, and coding activities might generate in a similar context. On Section II.B we discuss how these parameters were filled for our instances. Each dependence in WPDEP represents a work package ($wp_i$) whose development can only start after the requirements for a second work package ($wp_j$) are known.

From these work packages, we develop a project schedule. Our schedule is a partial order of software development activities on which four types of activities are created for each work package: requirement analysis, design, coding, and testing. A testing activity is always preceded by a coding activity, which is preceded by a design activity, which comes after a requirement analysis activity. These dependencies represent the basic workflow of modeling the problem, designing the solution, implementing its source-code, and testing its features. The schedule also includes dependencies amongst work packages: the requirements analysis activity for work package $wp_i$ cannot start before the requirements analysis activity for $wp_j$ is concluded for all $(wp_i, wp_j)$ pairs in WPDEP.

A schedule is an acyclic directed graph $<$AC, ACDEP$>$ consisting of a node set AC=\{$a_1, a_2, ..., a_n$\} of activities and an edge set ACDEP=\{($a_i, a_j$): i$\not=$j,1$\leq$i$\leq$m,1$\leq$j$\leq$m\} of finish-start (FS) dependencies amongst activities. A FS dependency A$\rightarrow$B indicates that work on activity B can only start when activity A is finished. Each activity is classified as either requirement analysis, design, coding, or testing. Analysis, design, and coding activities are described by the expected effort required to produce their results and the expected number of errors they might generate if performed by a mean developer in a regular environment. Testing activities are described by the average effort required to identify and correct an error, which is collected from information data in work packages.

We consider three objectives for overtime planning: the number of overtime hours (NOH), project makespan (MKS), and overall project cost (CST). OPP is formulated as a three objective decision problem in which makespan, cost, and overtime hours are conflicting objectives to be minimized. A candidate solution for the OPP is a sequence of numbers that represents the amount of daily overtime hours to be spent on each activity comprising the schedule. Our formulation is based on a 40-hours work week, accepts multiples of 30 minutes for overtime work, and limits the maximum amount of daily work to 12-hours turns, according to Brazilian law. Thus, for any given activity we may assign zero overtime, 30 minutes of overtime, one hour, one hour and 30 minutes, and so on up to four hours of daily overtime. Our first objective, NOH, is calculated as the sum of the amount of overtime hours assigned for each activity in the schedule.

The $<$ACT, ACDEP$>$ graph is comprised of a set of paths $\Pi$ denoting the precedence-respecting sequences of activities that must be executed to complete the project. Makespan is given by the longest path in $\Pi$ (critical path) and is calculated by Equation (1). The duration of an activity is calculated by a simulator according to the characteristics of the activity and its execution setup (including the number of daily overtime hours assigned to it). The simulator calculates the number of errors produced by each development activity and propagates these errors to succeeding activities. For testing activities, the simulator calculates the effort required to identify and correct all errors received from former activities.

\setlength\abovedisplayshortskip{0pt}
\setlength\belowdisplayshortskip{0pt}
\begin{small}
\begin{equation}
MKS = MAX_{p\in\Pi} (\sum_{a_{i} \in p} Duration (a_{i}))
\end{equation}
\end{small}

@@

\subsection{Simulation Mechanism}
Simulation is the imitation of the behavior presented by a real-world process or system over time \cite{Banks:2000}. It supports predicting and understanding the behavior of complex systems, containing hundreds or thousands of interconnected components. Simulation is useful because the number of components and the complexity of the data and control exchange among them usually hampers our ability to interpret system behavior by intuition or analytical reasoning.

@@

To overcome some of these problems, we have designed an object-oriented, continuous-time simulator to model the effects of overtime in software projects and predict project behavior (cost and makespan) accordingly. The simulator is based on three constructs: {\it simulation objects}, {\it resources}, and {\it scenarios}. Instead of creating a specific syntax to represent these elements, we relied on the Java programming language to describe and execute them. Designing the simulator in Java also simplified its integration to the heuristic search framework we have used in our experimental study.

{\it Simulation objects} are the elements whose behavior we want to simulate. They represent the nodes of a directed and acyclic graph whose edges denote precedence dependencies amongst simulation objects. They are initialized before the simulation starts and thereafter kept dormant until all their dependencies have been processed. Then, they become alive and start participating in the simulation until signaling their conclusion to the simulator. Simulation ends when all objects presented to the simulator are finished.

Time is managed in a continuous fashion. Simulation starts at time zero and is comprised of consecutive {\it simulation steps}, each advancing the simulator's clock by an infinitesimal increment. At each step, all alive simulation objects are allowed to update their internal state. Simulation objects are Java objects that implement an interface defining a protocol through which they are controlled by the simulator. Their state is defined by the relevant characteristics of the real-world element they represent in the simulation. An object's state is kept in its instance variables. State adjustment operations are coded in methods activated by the simulator: {\it init()}, to initialize the object at time zero; {\it start()}, called when all dependencies for an object are finished; {\it step()}, called at each simulation step; and {\it finish()}, called when the object signals its conclusion. The behavior of an object is represented by the history of changes observed in its internal state during the simulation.

During a simulation step, objects may use resources. A {\it resource} has a limited amount of materials or capabilities that can be consumed by simulation objects. These materials or capabilities are refilled before each simulation step. An object may request part of the resource's stock and adjust its internal state accordingly to its availability. The simulator controls the usage of limited resources by different simulation objects running in parallel. To model a software project, simulation objects represent an activity network with FS dependencies. The developers available to work on activities are resources whose amount of work is limited to a certain number of hours per day and is renewed in a daily basis. Developers are assigned to activities and if two or more activities executing in parallel share the same developer, only one of them will use the developer's capabilities at a time.

Scenarios are used to represent complementary behavior that may be imposed to certain simulation objects. A scenario is written to influence a certain type of simulation object and only participates in the simulation if activated on one or more objects of that type. Scenarios can query and update the internal state of their associated objects before and after each event on which the objects update their internal state ({\it start}, {\it step}, {\it finish}). Scenarios can be used to represent managerial decisions imposed to a subset of a software project's activities. They are reusable model components that modify the behavior of their related simulation object according to the policy they represent. For the purpose of the OPP, a scenario was developed to represent the effects of overtime while a base model describes the expected behavior for a project, its activities and developers as if uninfluenced by managerial policies and decisions (such as overtime allocation).

\subsection{The Base Model}
The base simulation model is comprised of activities and developers. Activities are simulation objects linked to each other by precedence dependencies. An activity is available for simulation as soon as its precedent activities are concluded. Developers are resources who can spend a limited amount of work on each simulation step. They are characterized by their productivity and error generation rates.

Activities are separated into development and testing activities. Development activities are described by an expected amount of work to be performed (in developer-hours) and an amount of errors expected to be generated while this work is performed. The duration of a given activity depends on the amount of work to be performed and the productivity of the developer assigned to it. We calibrated the model using data from Jones \cite{Jones:2000}. Given a work package measured in function-points (FP), Jones \cite{Jones:2000} presents average distributions of effort required for each development phase. For instance, for projects of about 1,000 function-points, requirement analysis activities consume 7\% of project effort, design activities consume about 10\%, coding activities consume about 30\%, and testing consumes 30\%. The remaining effort is related to management, documentation, and other activities. In our model, this effort is proportionally distributed among the analysis, design, coding, and testing activities. Given this distribution, the size of a work package, and an average productivity of 27.8 function-points per developer-month \cite{Jones:2000}, we calculate the expected duration for each activity.

Error dynamics is trickier. When an activity starts, it collects all errors produced by its former activities (if any) carrying them toward the end product. Abdel-Hamid and Madnick \cite{Abdel-Hamid:1991} present a model for error reproduction which states that errors inherited by an activity are not only forwarded to succeeding activities, but they contribute to the generation of new errors. Jones \cite{Jones:2000} suggests that the average number of defects introduced by each kind of activity is 1 error/FP in requirement analysis, 1.25 errors/FP in design, and 1.75 errors/FP in coding activities. Based on these data, our model assumes that each activity introduces 1 new error/FP, design activities regenerate 25\% of errors it receives from analysis, and coding regenerates 33\% of the errors received from design. These regeneration rates lead to the mark of 4.0 errors/FP on average after the implementation and are compatible with the former model \cite{Abdel-Hamid:1991} and source \cite{Jones:2000}.

Testing activities receive the results of development activities and aim at identifying and correcting the errors introduced in the product. These activities are characterized by the average effort required to identify and correct an error. Their duration depends on this parameter along with the assigned developer’s productivity. Given the amount of work expected to be invested in the testing phase \cite{Jones:2000} (i.e., 30\% of total effort for 1,000 FP projects) and the expected number of errors reaching this activity in a regular situation (4 errors/FP), we calculate the average time required to identify and correct each error and use this value on testing activities.

\subsection{A Scenario for Overtime}

The scenario used in our simulation describes the dynamics of working overtime and its influence on developer's productivity and error generation rates. If a developer is expected to work more than 8 hours per day in a given activity, his productivity will be increased proportionally to the extra number of work hours, reaching up to a 50\% increase if working to the upper limit of 12 hours per day. However, according to deMarco's law \cite{demarco:1999}, a developer working more than the regular daily hours may produce software below the quality standard. The more hours a developer works, the more tired he becomes. Tired developers become less careful to the work they are performing and, consequently, they tend to introduce more errors in the software. We have used data from \cite{Abdel-Hamid:1991}  to estimate the increase in the number of errors produced according to the number of hours worked by a developer. Such data suggest that software quality tends to deteriorate if developers work more than 9 hours/day and that a developer working 12 hours/day commits 50\% more errors than while working on a regular shift.

\section{Solution Approach}
Our solution uses the NSGAII algorithm for optimization along with the simulator described in Section II. NSGAII \cite{deb:2002} is a multi-objective genetic algorithm based on a ranking procedure which classifies candidate solutions according to their dominance. A candidate solution is described by the values calculated for the objectives of interest under that solution. In our case, these objectives are the number of overtime hours, project makespan, and project cost, all to be minimized. A solution A is said to dominate a solution B if the values of all objectives in A are no greater than the respective objectives in B and at least one objective is smaller in A than in B. Non-dominated solutions are assigned a rank of 1; solutions dominated only by non-dominated solutions are assigned a rank of 2; solutions dominated only by the former are assigned a rank of 3, and so on. The algorithm evolves a population over a number of generations, applying crossover, mutation, and selection upon candidate solutions. The selection process prioritizes low-ranking solutions and, when a subset of solutions having the same rank must be selected, a density measure allows selecting candidate solutions covering the search space as uniformly as possible.

As any heuristic algorithm, NSGAII has components that must be specified for a given problem. Three components are discussed below, while population size and stop criteria were subject of analysis and are discussed in the next section.

\noindent
\textbf{Representation}. A candidate solution for OPP is a sequence of m integer numbers, m being the number of activities. Each number in the solution pertains to the [0..8] integer interval and represents a multiple of 30-minutes overtime for a given activity. Thus, if the number 4 is assigned to activity A, A will receive 1.5 hours of overtime work. Each individual solution being evolved by NSGAII is evaluated through simulation. The simulator activates the overtime scenario upon each activity in the schedule according to the number of overtime hours assigned to that activity in the solution. Then, the simulator calculates the number of overtime hours, project makespan, and project cost for that solution.

\noindent
\textbf{Crossover}. We reused the crossover operator proposed by Ferrucci et al. \cite{Ferrucci:2013}. This operator was adopted because it was shown to outperform the one-point crossover operator commonly used in NSGAII for their formulation for the overtime planning problem. Instead of exchanging genetic material of two parents P$_{1}$ and P$_{2}$ after a point of cut C $\in$ [1..m], the resulting offspring O$_{1}$ and O$_{2}$ produced by this operator are defined by the equations below (given a random variable p with uniform distribution). The operator is executed with 50\% probability; otherwise, the parents are copied to the next generation. Parents are selected by binary tournament.

\setlength\abovedisplayskip{0pt}
\setlength\belowdisplayshortskip{0pt}
\begin{small}
\begin{equation}
O_{1}(g)=
\left \{
\begin{array}{ll}
P_{1}(g) & 1\leq g < C \\
max(P_{1}(g), P_{2}(g)) & C\leq g\leq m,p < 0.5 \\
min(P_{1}(g), P_{2}(g)) & C\leq g\leq m,p \geq 0.5 \\
\end{array}
\right.
\end{equation}

\setlength\abovedisplayshortskip{0pt}
\setlength\belowdisplayshortskip{0pt}
\small
\begin{equation}
O_{2}(g)=
\left \{
\begin{array}{ll}
P_{2}(g)& 1\leq g < C \\
P_{1}(g)+P_{2}(g) & C\leq g\leq m\\
\end{array}
\right.
\end{equation}
\end{small}

\noindent
\textbf{Mutation}. As in Ferrucci et al. \cite{Ferrucci:2013}, mutations are executed for all crossover offspring. Each gene of an offspring O is changed with 30\% probability, being assigned a new value in the [0..8] integer interval.

\section{Experimental Design}
This section explains the design of our empirical study, presenting research questions, the instances selected to address these questions, and the measurements used to compare solutions found under different configurations and algorithms.
\subsection{Research Questions}

In this section we present the research questions addressed by our study. They were designed to provide evidence on the validity, competitiveness, and usefulness of our approach.

\noindent
\textbf{RQ1 (Validation):} How does NSGAII perform if compared to random search on finding solutions for OPP? This is a "sanity check" to determine if our formulation allows a systematic search to outperform random search. If such cannot be verified, either the formulation is poorly designed, parameters are not adequate, or the problem is simple enough to allow a non-systematic procedure to find good solution.

\noindent
\textbf{RQ2 (Competitiveness):} How does NSGAII perform if compared to currently used overtime planning approaches? Besides beating random search, the proposed approach has to outperform current management practices related to overtime planning. In this sense, we compare the results produced by NSGAII with overtime planning practices used in the industry.

\noindent
\textbf{RQ3 (Usefulness):} Do results of overtime planning change if we consider the dynamics of error generation and the loss of quality caused by overtime? To address this issue, we compare the results produced by NSGAII with and without considering an increase in error generation due to overtime. Different results denote that a manager may pick a distinct overtime planning strategy if error dynamics are not considered.
